Reconstructing contact network parameters from viral phylogenies

### 2016-07-06 doc/paper/paper.tex
Add paper and tex

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage[backend=biber]{biblatex}
\usepackage{fullpage}
\usepackage{array}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{cleveref}

\addbibresource{papers.bib}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\onehalfspacing

\newcommand{\tablepath}{../tables}
\graphicspath{{../figures/}}

\title{Reconstructing contact network parameters from viral phylogenies}
\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
\abstract{\input{abstract}}

\maketitle

\section{Introduction}

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, sexual or in some
cases blood-to-blood contact is required. The population together with the set
of links between individuals along which transmission can occur is called the
contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
structure of the contact network underlying an epidemic can profoundly impact
the speed and pattern of the epidemic's expansion. Network structure can
influence the prevalence trajectory~\autocite{o2011contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates of
quantities such as effective population size~\autocite{goodreau2006assessing}.
From a public health perspective, contact networks have been explored as tools
for curtailing epidemic spread, by way of interventions targeted to
well-connected nodes~\autocite{wang2015targeting}. True contact networks are a
challenging type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the inter-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2012estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
suggesting that such statistical inference might be
possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale, colgate1989risk} have found that these networks tend to
have a degree distribution which follows a power law \autocite[although there has
been some disagreement, see][]{handcock2004likelihood}. That is, the number of
nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
$\gamma$. These networks are also creferred to as ``scale-free''. One process by
which scale-free networks can be generated is pcreferential attachment, where
nodes with a high number of contacts attract new connections at an elevated
rate. The first contact network model incorporating pcreferential attachment was
introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
Barab\'asi-Albert (BA) model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. In the most commonly studied
formulation, these new edges are joined to existing nodes of degree $k$ with
probability proportional to $k$, so that nodes of high degree tend to attract
more connections. \citeauthor{barabasi1999emergence} suggested an extension
where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
some non-negative constant $\alpha$, and we use this extension here.

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a Bayesian
approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
equations describing the spread of a susceptible-infected (SI) epidemic on
static and dynamic contact networks with several degree distributions, which
could in principle be used for inference if observed incidence trajectories
were available. \textcite{brown2011transmission} analysed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV-infected men who have
sex with men (MSM) in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection. The authors found that a Waring distribution,
which is produced by a more sophisticated pcreferential attachment model, was a
good fit to their estimated network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
which the normalizing constants cancel out. Unfortunately, it is generally
difficult to explicitly calculate the likelihood of an observed transmission
tree under a contact network model, even up to a normalizing constant. To do
so, it would be necessary to integrate over all possible networks, and also
over all possible labellings of the internal nodes of the transmission tree.
While it is not known (to us) whether such integration is tractable, a simpler
alternative is offered by likelihood-free methods, namely approximate Bayesian
computation (ABC). ABC leverages the fact that, although calculating the
likelihood may be impractical, generating simulated datasets according to a
model is often straightforward. If our model fits the data well, the simulated
data it produces should be similar to the observed data. More formally, if $D$
is the observed data, the posterior distribution $f(\theta \mid D)$ on model
parameters $\theta$ is replaced as the target of statistical inference by
$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.

Here, we develop a method using kernel-ABC to estimate the parameters of
contact network models from observed phylogenetic data. The distance function
we use is the tree kernel developed by \textcite{poon2013mapping}, which computes
a weighted dot product of the trees' representations in the space of all
possible subset trees. We apply the method to investigate the parameters of the
BA network model on a variety of simulated and real datasets. Our results show
that some network parameters can be inferred with reasonable accuracy, while
others have a minimal detectable impact on tree shape and thecrefore cannot be
estimated accurately. We also find that these parameters can vary considerably
between real epidemics from different settings.

\section{Methods}

\subsection{\textit{Netabc}: phylogenetic inference of contact network
parameters with kernel-ABC}

We have developed a kernel-ABC-based method to perform statistical inference of
contact network parameters from a transmission tree estimated from an observed
viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
algorithm for ABC developed by \textcite{del2012adaptive}. The SMC algorithm
keeps track of a population of parameter ``particles'', which are initially
sampled from the parameters' joint prior distribution. Several datasets are
simulated under the model of interest for each of the particles. In this case,
the datasets are transmission trees, which are generated by a two-step process.
First, a contact network is simulated according to the network model being fit.
Second, a transmission tree is simulated over that network with a Gillespie
simulation algorithm~\autocite{gillespie1976general}, in the same fashion as
several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
leventhal2012inferring}. The particles are weighted according to the similarity
between their associated simulated trees and the observed tree. To quantify
this similarity, we used the tree kernel developed by
\textcite{poon2013mapping}. Particles are iteratively perturbed by applying a
Metropolis-Hastings kernel and, if the move is accepted, simulating new
datasets under the new parameters. When a particle's weight drops to zero,
because its simulated trees are too dissimilar to the observed tree, the
particle is dropped from the population, and eventually replaced by a resampled
particle with a higher weight. As the algorithm progresses, the population
converges to a Monte Carlo approximation of the ABC target distribution, which
is assumed to approximate the desired posterior~\autocite{del2012adaptive,
sunnaaker2013approximate}. A computer program implementing our method is freely
available at \url{https://github.com/rmcclosk/netabc} (last accessed April 3,
2016).

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
demonstrate their method (originally used by~\textcite{sisson2007sequential}).
We were able to obtain a close approximation to the function (see
\cref{fig:smctest}), and attained the stopping condition used by the authors in
a comparable number of steps.

Nodes in our networks followed simple SI dynamics, meaning that they became
infected at a rate proportional to their number of infected neighbours, and
never recovered. For all analyses, the transmission trees' branch lengths were
scaled by dividing by their mean. We used the \textit{igraph} library's
implementation of the BA model~\autocite{csardi2006igraph} to generate the
graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
a local computer cluster.

\subsection{Kernel classifiers}

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the BA model had an effect on tree shape. 100
networks were simulated under each of three different values of $\alpha$: 0.5,
1.0, and 1.5 (300 networks total). The other parameters were fixed to the
following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
500 tips was simulated over each network (300 transmission trees total). The
300 trees were compared pairwise with the tree kernel to form a $300 \times
300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
factor''), and $\sigma$ (the ``radial basis function
variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
two-fold cross-validations.

Three similar experiments were performed for the other BA model parameters (one
experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
1. Thus, we performed a total of four kSVR cross-validations, one for each of
the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
differing numbers of tips (100, 500, and 1000) and in epidemics of differing
size (500, 1000, and 2000). The combination of the number of sampled
individuals (\textit{i.e.} the number of tips) and the epidemic size
(\textit{i.e.} $I$) will be creferred to as an ``epidemic scenario''. When
evaluating the classifier for $I$, we did not consider trees with 1000 tips,
because one of the tested $I$ values was 500, and the number of tips cannot be
larger than $I$.

For each of the four parameters, we also tested a linear regression against
Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.

\subsection{ABC simulations}

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
per combination = 72 trees total. The adaptive ABC algorithm was applied to
each tree with these priors: $m \sim$ DiscreteUniform(1, 5), $\alpha \sim$
Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Proposals for $\alpha$, $N$, and
$I$ were Gaussian, while proposals for $m$ were Poisson. Following
\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, the variance of
all proposals was equal to the empirical variance of the particles.

The algorithm was run with 1000 particles, 5 simulated datasets per particle,
and the ``quality'' parameter controlling the decay rate of the tolerance
$\varepsilon$ set to 0.95. We used the same stopping criterion as
\citeauthor{del2012adaptive}, namely when the MCMC acceptance rate dropped
below 1.5\%. Point estimates for the parameters were obtained by taking the
highest point of an estimated kernel density on the final set of particles,
calculated using the \textit{density} function with the default parameters in
\textit{R}. Highest posterior density (HPD) intervals were calculated with the
\textit{HPDinterval} function from the \textit{R} package
\textit{coda}~\autocite{plummer2006coda}.

Two further analyses were performed to address potential sources of error. To
evaluate the effect of model misspecification in the case of heterogeneity
among nodes, we generated a network where half the nodes were attached with
power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
investigate the effects of potential sampling bias, we simulated a transmission
tree where the tips were sampled in a peer-driven fashion, rather than at
random. That is, the probability to sample a node was twice as high if any of
that node's network peers had already been sampled. The parameters of this
network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.

\subsection{Investigation of published data}

We applied our kernel-ABC method to several published HIV datasets. Because the
BA model generates networks with a single connected component, we
specifically searched for datasets which originated from existing clusters,
either phylogenetically or geographically defined. Characteristics of the
datasets we investigated are given in \cref{tab:data}.

\begin{table*}[!t]
  \centering
  \input{\tablepath/realdata}
  \caption{Characteristics of published datasets investigated with kernel-ABC.
  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
  \label{tab:data}
\end{table*}

We downloaded all sequences associated with each study from GenBank. For the
\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
creference sequence (Genbank accession number K03455) and the hypervariable
regions were clipped out with \textit{BioPython} version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\textit{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments were
manually inspected with \textit{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \textit{FastTree2} version
2.1.7 with the generalized time-reversible model. Transmission trees were
estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
larger than the others, containing 1265 and 1299 sequences respectively. To
ensure that the analyses were comparable, we reduced these to a number of
sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
described previously~\autocite{poon2015impact}. Only sequences within this
cluster were carried forward. For the \textcite{novitsky2014impact} data, no
large clusters were detected using the same cutoff, so we analysed a subtree of
size 180 chosen arbitrarily.

For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
value $m = 1$ produces networks with no cycles, which we considered fairly
implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
the SMC algorithm were the same as used for the simulation experiments.

\section{Results}

\subsection{Kernel classifiers}

>>>>

\usepackage{alltt}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{fullpage}
\usepackage{array}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{cleveref}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{amsmath}

\addbibresource{papers.bib}

% beginning of diff stuff
\usepackage{twoopt}

\let\mref\cref
\let\Mref\Cref
\let\mtextcite\textcite
\let\mautocite\autocite
\let\mciteauthor\citeauthor
\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
\renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
\renewcommandtwoopt{\autocite}[3][][]{\mbox{\mautocite[#1][#2]{#3}}}
\renewcommand{\citeauthor}[1]{\mbox{\mciteauthor{#1}}}

\newcommand{\add}[1]{\color{blue} \uline{#1} \color{black}}
\newcommand{\del}[1]{\color{red} \sout{#1} \color{black}}
% end of diff stuff

\newcommand{\tablepath}{../tables}
\graphicspath{{../figures/}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\onehalfspacing

\title{Reconstructing contact network parameters from viral phylogenies}
\author{Rosemary M. McCloskey$^1$, Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$ \\
\small $^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ \small $^2$ Department of Medicine, University of British Columbia, Vancouver, Canada}

\maketitle
\abstract{\input{abstract}}

\section*{Introduction}

When an infectious disease spreads through a population, transmissions are
generally more likely to occur between certain pairs of individuals. Such pairs
must have a particular mode of contact with one another, which varies with the
mode of transmission of the disease. For airborne pathogens, physical proximity
may be sufficient, while for sexually transmitted diseases, sexual or in some
cases blood-to-blood contact is required. The population together with the set
of links between individuals along which transmission can occur is called the
contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
structure of the contact network underlying an epidemic can profoundly impact
the speed and pattern of the epidemic's expansion. Network structure can
influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
the estimates of quantities such as effective viral population
size~\autocite{goodreau2006assessing}.  From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical, eames2015six}.

Viral sequence data, on the other hand, has become \del{relatively inexpensive
and straightforward to collect on a population level.} \add{easier to collect
as the cost of sequencing has declined. In the case of HIV, genotyping has
become part of routine clinical care in several health regions.} Due to the
high mutation rate of RNA viruses, epidemiological processes impact the course
of viral evolution, thereby shaping the inter-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2012estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, although
sophisticated methods have been developed for fitting complex population
genetic models to phylogenies~\autocite{volz2012complex,
rasmussen2014phylodynamic}, inference of structural network parameters has to
date been limited. However, it has been shown that network structure has a
tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
villandre2016assessment}, suggesting that such statistical inference might be
possible~\autocite{welch2011statistical}. \add{In the context of networks,
sequence data have the advantage of being objective, in that they are not
affected by misreporting. However, just as with survey data, it is important to
collect a representative sample from the population to perform accurate
inference~\autocite{novitsky2014impact}.}

Survey-based studies of sexual networks~\autocite{colgate1989risk,
liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
clemenccon2015statistical} have found that these networks tend to have a degree
distribution which follows a power law \autocite[although there has been some
disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
networks are also referred to as
``scale-free''~\autocite{barabasi1999emergence}. One process by which
scale-free networks can be generated is preferential attachment, where nodes
with a high number of contacts attract new connections at an elevated rate. The
first contact network model incorporating preferential attachment was
introduced by \textcite{barabasi1999emergence}, and is now referred to as the
Barab\'asi-Albert (BA) model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. In the most commonly studied
formulation, these new edges are joined to existing nodes of degree $k$ with
probability proportional to $k$, so that nodes of high degree tend to attract
more connections. \citeauthor{barabasi1999emergence} suggested an extension
where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
some non-negative constant $\alpha$, and we use this extension in this work.
\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
for $\alpha < 1$, the distribution is a stretched exponential, while for
$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
nodes are connected to nearly every other node in the
graph~\autocite{krapivsky2000connectivity}.}

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a Bayesian
approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
equations describing the spread of a susceptible-infected (SI) epidemic on
static and dynamic contact networks with several degree distributions, which
could in principle be used for inference if observed incidence trajectories
were available. \textcite{brown2011transmission} analyzed the degree distribution
of an approximate transmission network, estimated based on genetic similarity
and estimated times of infection, relating 60\% of HIV-infected men who have
sex with men (MSM) in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection. The authors found that a Waring distribution,
which is produced by a more sophisticated preferential attachment model, was a
good fit to their estimated network.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
which the normalizing constants cancel out. Unfortunately, it is generally
difficult to explicitly calculate the likelihood of an observed transmission
tree under a contact network model, even up to a normalizing constant. To do
so, it would be necessary to integrate over all possible networks, and also
over all possible labellings of the internal nodes of the transmission tree.
While it is not known (to us) whether such integration is tractable, a simpler
alternative is offered by likelihood-free methods, namely approximate Bayesian
computation (ABC)~\autocite{tavare1997inferring, beaumont2002approximate}. ABC
leverages the fact that, although calculating the likelihood may be
impractical, generating simulated datasets according to a model is often
straightforward. If our model fits the data well, the simulated data it
produces should be similar to the observed data. More formally, if $D$ is the
observed data, the posterior distribution $f(\theta \mid D)$ on model
parameters $\theta$ is replaced as the target of statistical inference by
$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
the specific case when $\rho$ is a kernel function, the approach is known as
ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
demonstrated that taking $\rho$ to be a well-chosen kernel function can produce
a more accurate ABC approximation than the typical choice of a difference of
summary statistics.}

Here, we develop a method using ABC to estimate the parameters of contact
network models from observed phylogenetic data. The distance function we use is
the tree kernel developed by \textcite{poon2013mapping}, which computes a
weighted dot product of the trees' representations in the space of all possible
subset trees. We apply the method to investigate the parameters of the BA
network model on a variety of simulated and real datasets. Our results show
that some network parameters can be inferred with reasonable accuracy, while
others \del{have a minimal detectable impact on tree shape and therefore cannot
be estimated accurately}\add{are weakly- or non-identifiable with ABC}. We also
find that these parameters can vary considerably between real epidemics from
different settings.

\section*{Methods}

\subsection*{\textit{Netabc}: phylogenetic inference of contact network
parameters with ABC}

We have developed an ABC-based method to perform statistical inference of
contact network parameters from a transmission tree estimated from an observed
viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
algorithm for ABC developed by \textcite{del2012adaptive}. The SMC algorithm
keeps track of a population of parameter ``particles'', which are initially
sampled from the parameters' joint prior distribution. Several datasets are
simulated under the model of interest for each of the particles. In this case,
the datasets are transmission trees, which are generated by a two-step process.
First, a contact network is simulated according to the network model being fit.
Second, a transmission tree is simulated over that network with a Gillespie
simulation algorithm~\autocite{gillespie1976general}, in the same fashion as
several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
leventhal2012inferring}. \add{Tips of the simulated transmission tree are
randomly removed until the simulated tree has the same number of tips as the
input tree.} The particles are weighted according to the similarity between
their associated simulated trees and the observed tree. To quantify this
similarity, we used the tree kernel developed by \textcite{poon2013mapping}.
Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
and, if the move is accepted, simulating new datasets under the new parameters.
When a particle's weight drops to zero,
because its simulated trees are too dissimilar to the observed tree, the
particle is dropped from the population, and eventually replaced by a resampled
particle with a higher weight. As the algorithm progresses, the population
converges to a Monte Carlo approximation of the ABC target distribution, which
is assumed to approximate the desired posterior~\autocite{del2012adaptive,
sunnaaker2013approximate}.

\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
beaumont2009adaptive}, the user is required to specify a decreasing sequence of
tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
associated simulated datasets within distance $\varepsilon_i$ of the observed
data are removed from the population. In the adaptive version of
\textcite{del2012adaptive}, the sequence of tolerances is determined
automatically by fixing the decay rate of the population's expected sample size
(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
confusion with the preferential attachment power parameter of the BA model.}

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the imbalance of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
demonstrate their method (originally used by~\textcite{sisson2007sequential}).
We were able to obtain a close approximation to the function (see
\cref{fig:smctest}), and attained the stopping condition used by the authors in
a comparable number of steps.

Nodes in our networks followed simple SI dynamics, meaning that they became
infected at a rate proportional to their number of infected neighbours, and
never recovered. For all analyses, the transmission trees' branch lengths were
scaled by dividing by their mean. We used the \textit{igraph} library's
implementation of the BA model~\autocite{csardi2006igraph} to generate the
graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
a local computer cluster. \add{A computer program implementing our method is
freely available at \url{https://github.com/rmcclosk/netabc} (last accessed
\today).}

\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}

\add{We considered four parameters related to the BA model, denoted $N$, $m$,
$alpha$, and $I$. The first three of these parameterize the network structure,
while $I$ is related to the simulation of transmission trees over the network.
However, we will refer to all four as BA parameters. $N$ denotes the total
number of nodes in the network, or equivalently, susceptible individuals in the
population. $m$ is the number of new undirected edges added for each new
vertex, or equivalently one-half of the average degree. $\alpha$ is the power
of preferential attachment -- new nodes are attached to existing nodes of
degree $d$ with probability proportional to $d^{\alpha} + 1$. Finally, $I$ is
the number of infected individuals at the time when sampling occurs. The
$\alpha$ parameter is unitless, while $m$ has units of edges or connections per
vertex, and $N$ and $I$ both have units of nodes or individuals.}

\add{Before proceeding with a full validation of \textit{netabc} on simulated
data, we undertook an experiment designed to assess the identifiability of the
BA parameters. One parameter of the BA model was investigated at a time while
holding all others fixed, a strategy commonly used when performing sensitivity
analyses of mathematical models. This allowed us to perform a fast preliminary
analysis without dealing with the ``curse of dimensionality'' of the full
parameter space. We simulated trees under three different values of each
parameter, and asked how well we could tell the different trees apart. The
better we are able to distinguish the trees, the more identifiability we might
expect for the corresponding parameter when we attempt to estimate it with
ABC.}

\add{This experiment also had the secondary purpose of validating our choice of
the tree kernel as a distance measure in ABC. To tell the trees apart, we used
a classifier based on the tree kernel, but we also tested two other tree shape
statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
imbalance which not take branch lengths into account, considering only the
topology. The normalized lineages-through-time
\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
distributions of their branching times, and does not explicitly consider the
topology. Since the tree kernel incorporates both of these sources of
information, we expected it to outperform the other two statistics. Finally,
the tree kernel can be tuned by adjusting the values of the meta-parameters
$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
variance'', see \textcite{poon2013mapping}). $\lambda$ is used to penalize
ween large subset trees which tend to dominate the kernel score. When $\lambda
= 0$, all but the root branches of each subset tree are ignored, while when
$\lambda = 1$, no penalty is applied. $\sigma$ controls how strictly the notion
of similarity is applied to branch lengths. When $\sigma = 0$, branch lengths
must match exactly, while as $\sigma \to \infty$, branch lengths are not
considered at all.}

\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
To test whether the parameters of the BA model had a \add{measurable} effect on
tree shape, 100 networks were simulated under each of three different values of
$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
transmission tree with 500 tips was simulated over each network (300
transmission trees total). The 300 trees were compared pairwise with the tree
kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVM)
\add{from the kernel matrix} with the \textit{kernlab}
package~\autocite{zeileis2004kernlab}, and two ordinary SVMs \add{from the nLTT
matrix and Sackin's index vector} with the e1071
package~\autocite{meyer2015e1071}.\del{, and a linear regression from the
Sackin's index values.} The accuracy of each classifier was evaluated with 1000
two-fold cross validations \add{with equally-sized folds}.

Three similar experiments were performed for the other BA model parameters (one
experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
1. Thus, we performed a total of four cross-validations \add{for each
classifier}, one for each of the BA model parameters $\alpha$, $I$, $m$, and
$N$. We repeated these four cross-validations with different values of
$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
as well as on trees with differing numbers of tips (100, 500, and 1000).
\add{For the structural parameters $\alpha$, $m$, and $N$, the experiments were
repeated with three different fixed values of $I$ (500, 1000, and 2000).}
\del{and in epidemics of differing size (500, 1000, and 2000).} The combination
of the number of sampled individuals (\textit{i.e.} the number of tips) and the
epidemic size (\textit{i.e.} $I$) will be referred to as an ``epidemic
scenario''. When evaluating the classifier for $I$, we did not consider trees
with 1000 tips, because one of the tested $I$ values was 500, and the number of
tips cannot be larger than $I$.

\del{For each of the four parameters, we also tested a linear regression against
Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the normalized
lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}

\subsection*{ABC simulations}
